---
title: "Exploratory Analysis Course Project 1"
author: "Angela Moturi"
date: '2022-08-20'
output: html_document
---
Loading and Pre-Processing Data
```{r load data}
# Loading packages
library(ggplot2)
library(dplyr)

# Unzipping the file and reading it
setwd("C:/Users/Amutori/Downloads")
activity <- read.csv("activity.csv")

# Setting date format to get the weekdays and weekends
activity$date <- as.POSIXct(activity$date, "%Y%m%d")

# Obtain days on the dataset
day <- weekdays(activity$date)

# Combining the dataset with the weekday of the dates
activity <- cbind(activity, day)
str(activity)
``` 
Question 1 -Mean total number of steps taken per day?
```{r mean steps}
# Calculating total steps taken on a day
activityTotalSteps <- with(activity, aggregate(steps, by = list(date), sum, na.rm = TRUE))
# Changing col names
names(activityTotalSteps) <- c("Date", "Steps")

# Converting the data set into a data frame to be able to use ggplot2
totalStepsdf <- data.frame(activityTotalSteps)

# Plotting a histogram using ggplot2
g <- ggplot(totalStepsdf, aes(x = Steps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "green", col = "black") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Per Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken on a Day")

print(g)
```
The mean of the total number of steps taken per day is:
```{r mean steps taken}
mean(activityTotalSteps$Steps)
```
The median of the total number of steps taken per day is:
```{r median steps taken}
median(activityTotalSteps$Steps)

```

Question 2 - average daily activity pattern
```{r mean steps by interval}
# Calculating the average number of steps taken, averaged across all days by 5-min intervals.
averageDailyActivity <- aggregate(activity$steps, by = list(activity$interval), 
                                  FUN = mean, na.rm = TRUE)
# Changing col names
names(averageDailyActivity) <- c("Interval", "Mean")

# Converting the data set into a dataframe
averageActivitydf <- data.frame(averageDailyActivity)

# Plotting on ggplot2
da <- ggplot(averageActivitydf, mapping = aes(Interval, Mean)) + 
  geom_line(col = "red") +
  xlab("Interval") + 
  ylab("Average Number of Steps") + 
  ggtitle("Average Number of Steps Per Interval") 
  
print(da)
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max}
averageDailyActivity[which.max(averageDailyActivity$Mean), ]$Interval

```
Question 3 - Imputing Missing Values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs.
```{r imputation}
sum(is.na(activity$steps))

```
Strategy for filling is using the mean for the 5-minute interval
```{r imputation with mean}
# Matching the mean of daily activity with the missing values
imputedSteps <- averageDailyActivity$Mean[match(activity$interval, averageDailyActivity$Interval)]
```
Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r imputed dataset}
# Transforming steps in activity if they were missing values with the filled values from above.
activityImputed <- transform(activity, 
                             steps = ifelse(is.na(activity$steps), yes = imputedSteps, no = activity$steps))

# Forming the new dataset with the imputed missing values.
totalActivityImputed <- aggregate(steps ~ date, activityImputed, sum)

# Changing col names
names(totalActivityImputed) <- c("date", "dailySteps")

#Check if new data still has any missing values

sum(is.na(totalActivityImputed$dailySteps))

```
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r plot1}
# Converting the data set into a data frame to be able to use ggplot2
totalImputedStepsdf <- data.frame(totalActivityImputed)

# Plotting a histogram using ggplot2
p <- ggplot(totalImputedStepsdf, aes(x = dailySteps)) + 
  geom_histogram(breaks = seq(0, 25000, by = 2500), fill = "purple", col = "black") + 
  ylim(0, 30) + 
  xlab("Total Steps Taken Each Day") + 
  ylab("Frequency") + 
  ggtitle("Total Number of Steps Taken per Day")

print(p)
```
The mean of the total number of steps taken per day is:
```{r mean steps 2}
mean(totalActivityImputed$dailySteps)

```
The median of the total number of steps taken per day is:
```{r median steps 2}
median(totalActivityImputed$dailySteps)

```
Question 4 - Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r factor}
# Updating format of the dates
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))

# Creating a function that distinguises weekdays from weekends
activity$dayType <- sapply(activity$date, function(x) {
  if(weekdays(x) == "Saturday" | weekdays(x) == "Sunday")
  {y <- "Weekend"}
  else {y <- "Weekday"}
  y
})
```
Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r plot2}
# Creating the data set that will be plotted
activityByDay <-  aggregate(steps ~ interval + dayType, activity, mean, na.rm = TRUE)

# Plotting using ggplot2
dayPlot <-  ggplot(activityByDay, aes(x = interval , y = steps, color = dayType)) + 
  geom_line() + ggtitle("Average Daily Steps by Day Type") + 
  xlab("Interval") + 
  ylab("Average Number of Steps") +
  facet_wrap(~dayType, ncol = 1, nrow=2) +
  scale_color_discrete(name = "Day Type") 
print(dayPlot) 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
